---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 5 - Due date 02/19/24"
author: "Julia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A05_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(readxl)
library(stats)
library(tseries)
library(ggplot2)
library(Kendall)
library(lubridate)
library(tidyverse)
library(cowplot)
```

## Decomposing Time Series

Consider the same data you used for A04 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2023 Monthly Energy Review.

```{r}
#Importing data set - using xlsx package
energy_data <- read_excel("~/Julia_Kagiliery_TSA_Sp24/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", 
    skip = 9)
energy_data <- energy_data[-1,]
```


### Q1

For this assignment you will work only with the following columns: Solar Energy Consumption and Wind Energy Consumption. Create a data frame structure with these two time series only and the Date column. Drop the rows with *Not Available* and convert the columns to numeric. You can use filtering to eliminate the initial rows or convert to numeric and then use the drop_na() function. If you are familiar with pipes for data wrangling, try using it!

```{r}
energy_data <- energy_data |> 
  select(c("Solar Energy Consumption", "Wind Energy Consumption", "Month"))

energy_data <- energy_data |>
   filter(
     `Solar Energy Consumption` != "Not Available",
     `Wind Energy Consumption` != "Not Available")

energy_data <- na.omit(energy_data)

energy_data <- energy_data |>
   mutate(
     `Solar Energy Consumption` = as.numeric(`Solar Energy Consumption`),
     `Wind Energy Consumption` = as.numeric(`Wind Energy Consumption`), 
     Month = ymd(Month)
   )
```

Chat GPT prompt for ymd(): how to use my() function for date conversions

```{r}
nobs=nrow(energy_data)
nvar=ncol(energy_data)
```

```{r}
energyTS <- ts(energy_data)
```

### Q2

Plot the Solar and Wind energy consumption over time using ggplot. Plot each series on a separate graph. No need to add legend. Add informative names to the y axis using `ylab()`. Explore the function `scale_x_date()` on ggplot and see if you can change the x axis to improve your plot. Hint: use `scale_x_date(date_breaks = "5 years", date_labels = "%Y")")`

```{r}
energy_data |>
ggplot(aes(x=Month, y= `Solar Energy Consumption`)) +
            geom_line(color="blue") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  ylab("Solar Energy Consumption (Trillion BTU)")+
  xlab("Year")
```

```{r}
energy_data |>
ggplot(aes(x=Month, y= `Wind Energy Consumption`)) +
            geom_line(color="red") +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    ylab("Wind Energy Consumption (Trillion BTU)")+
  xlab("Year")
```


### Q3

Now plot both series in the same graph, also using ggplot(). Use function `scale_color_manual()` to manually add a legend to ggplot. Make the solar energy consumption red and wind energy consumption blue. Add informative name to the y axis using `ylab("Energy Consumption)`. And use function `scale_x_date()` to set x axis breaks every 5 years.

```{r}
energy_data |>
  ggplot(aes(x = Month, color = factor("Wind"), y = `Wind Energy Consumption`)) +
  geom_line() +
  geom_line(aes(x = Month, color = factor("Solar"), y = `Solar Energy Consumption`)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  ylab("Energy Consumption") +
  xlab("Year") +
  scale_color_manual(name = "Energy Source", values = c("Wind" = "blue", "Solar" = "red"))
```
Code inspo: Chat GPT. Prompt: why isn't "scale_color_manual(values = colors)" adding a legend

## Decomposing the time series

The stats package has a function called decompose(). This function only take time series object. As the name says the decompose function will decompose your time series into three components: trend, seasonal and random. This is similar to what we did in the previous script, but in a more automated way. The random component is the time series without seasonal and trend component. 

```{r}
WindTS <- ts(energy_data$`Wind Energy Consumption`, start = c(1984,1), frequency = 12)
SolarTS <- ts(energy_data$`Solar Energy Consumption`, start = c(1984,1), frequency = 12)
```

```{r}
autoplot(WindTS)
```

```{r}
autoplot(SolarTS)
```


Additional info on `decompose()`.

1) You have two options: alternative and multiplicative. Multiplicative models exhibit a change in frequency over time.
2) The trend is not a straight line because it uses a moving average method to detect trend. 
3) The seasonal component of the time series is found by subtracting the trend component from the original data then grouping the results by month and averaging them. 
4) The random component, also referred to as the noise component, is composed of all the leftover signal which is not explained by the combination of the trend and seasonal component. 

### Q4

Transform wind and solar series into a time series object and apply the decompose function on them using the additive option, i.e., `decompose(ts_data, type = "additive")`. What can you say about the trend component? What about the random component? Does the random component look random? Or does it appear to still have some seasonality on it?

> Answer: The trend componet looks about how I would expect it: increasing. However, the random component does not look random at all; there certainly seems to be some seasonaility left. The random plots almost seem to imply that data before 2012 is one model and data after 2012 is a totally different one.  

```{r}
Q4WindTrend <- decompose(WindTS, type = "additive")
```

```{r}
autoplot(Q4WindTrend$seasonal)
```

```{r}
autoplot(Q4WindTrend$trend)
```

```{r}
autoplot(Q4WindTrend$random)
```

```{r}
Q4SolarTrend <- print(decompose(SolarTS, type = "additive"))
```

```{r}
autoplot(Q4SolarTrend$seasonal)
```

```{r}
autoplot(Q4SolarTrend$trend)
```

```{r}
autoplot(Q4SolarTrend$random)
```

### Q5

Use the decompose function again but now change the type of the seasonal component from additive to multiplicative. What happened to the random component this time?

> Answer: The random component looks better this time (expecially for wind) as there is more randomness to the plot and hence less evidence of seasonaility. There is still some indication that the timefram should be fractionated for models.

```{r}
Q5Wind <- decompose(WindTS, type = "multiplicative")
```

```{r}
autoplot(Q5Wind$seasonal)
```

```{r}
autoplot(Q5Wind$trend)
```

```{r}
autoplot(Q5Wind$random)
```

```{r}
Q5Solar <- decompose(SolarTS, type = "multiplicative")
```

```{r}
autoplot(Q5Solar$seasonal)
```

```{r}
autoplot(Q5Solar$trend)
```

```{r}
autoplot(Q5Solar$random)
```

```{r}
autoplot(WindTS)
```

```{r}
autoplot(SolarTS)
```

### Q6

When fitting a model to this data, do you think you need all the historical data? Think about the data from 90s and early 20s. Are there any information from those years we might need to forecast the next six months of Solar and/or Wind consumption. Explain your response.

> Answer: No, I do not think all of the historic data is neccessary. There have been many factors over time like increased environmental awareness, technology improvemnets as well as cost reductions in solar and wind technology all of whihc have fundamentally changed the way we consumer solar and wind energy. Hence, it is probably better to look at the last year or two to predict the next six months as the technology from 1 or 2 years ago is far similar to the systems we are using today than infomration/technology from 30 years ago.  

### Q7

Create a new time series object where historical data starts on January 2012. Hint: use `filter()` function so that you don't need to point to row numbers, .i.e, `filter(xxxx, year(Date) >= 2012 )`. Apply the decompose function `type=additive` to this new time series. Comment the results. Does the random component look random? Think about our discussion in class about seasonal components that depends on the level of the series. 

```{r}
FilterData <- energy_data|>
  filter(year(as.Date(Month, "%Y")) >= 2012)
```

```{r}
FilterDataTS <- ts(FilterData)
```

```{r}
FilterWind <-FilterData |>
  select(`Wind Energy Consumption`)

FilterSolar <-FilterData |>
  select(`Solar Energy Consumption`)
```

```{r}
TSFilterWind <- ts(FilterWind, start = c(2012, 1), frequency = 12)
TSFilterSolar <- ts(FilterSolar, start = c(2012, 1), frequency = 12)
```

```{r}
Q6Wind <- decompose(TSFilterWind, type = "additive")
```

```{r}
Q6Solar <- decompose(TSFilterSolar, type = "additive")
```

```{r}
autoplot(Q6Solar$random)
```

```{r}
autoplot(Q6Wind$random)
```

```{r}
autoplot(TSFilterSolar)
```

```{r}
autoplot(TSFilterWind)
```

> Answer: The random components do look much better here and are far more random; there are no distinguishable patterns that emerge year to year. Of course, this will never be completely ranomd as we expect some noise in everydata set but this version of the data is the best explained by the model so far.  

## Identify and Remove outliers

### Q8

Apply the `tsclean()` to both series from Q7. Did the function removed any outliers from the series? Hint: Use `autoplot()` to check if there is difference between cleaned series and original series.

> Answer:When considering substantial historical data in the data set, a ton of outliers are removed for solar and wind (you know this because the two trends are nit super imposed). There seem to be a variety of factors (like new technology and increased environmnetal awareness) that have changed the energy consumption in these two categories. The variability across time tells us that historic data may not be the best predictor of current and future uses. 

```{r}
CleanWindTS <-tsclean(WindTS)
```

```{r}
CleanSolarTS <- tsclean(SolarTS)
```

```{r}
autoplot(SolarTS, series="Solar Raw") +
  autolayer(CleanSolarTS, series="Solar Clean") +
  ylab("Energy Consumption") 
```

```{r}
autoplot(WindTS, series="Wind Raw") +
  autolayer(CleanWindTS, series="Wind Clean") +
  ylab("Energy Consumption") 
```

### Q9

Redo number Q8 but now with the time series you created on Q7, i.e., the series starting in 2014. Using what `autoplot()` again what happened now?Did the function removed any outliers from the series?
After 2012

> Answer: With the series including data only from 2012 on, there were very few outliers removed in the solar data (right around Janurary of 2022 and 2023) but otherwise no data was removed (the two series are almost completely superimposed). This seems to show that it is better/ more predicatble to look at data around the same time rather than using largerly historic data. 


```{r}
TSCWind <- tsclean(TSFilterWind)
TSCSolar <- tsclean(TSFilterSolar)
```

```{r}
autoplot(TSFilterSolar, series="Solar Raw") +
  autolayer(TSCSolar, series="Solar Clean") +
  ylab("Energy Consumption") 
```

```{r}
autoplot(TSFilterWind, series="Wind Raw") +
  autolayer(TSCWind, series="Wind Clean") +
  ylab("Energy Consumption") 
```

